### 2015-10-26 manuscript.Rnw
manuscript edits: now with spelling\!

The R programming language is a popular tool for  phylogenetics and comparative
analysis. The R packages \texttt{ape} \cite{Paradis2004}, \texttt{phylobase}
\cite{phylobase-0.8.0 }and \texttt{RNeXML} \cite{Boettiger2015} provide
functions to import and manipulate trees within R and save the results in
standard data formats. Additional packages including  \texttt{phytools}
\cite{Revell2012}, \texttt{geiger} \cite{Pennell2014-geiger} and \texttt{ggtree} 
\cite{Yu-ggtree} to allow users to analyse and visualise data in a 
phylogenetic context (see 
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list of phylogenetics packages in R). In addition to packages for
phylogentic analyses, a growing number of R packages allow users to query and
acess data from the web \hl{(rFISHBASE 10.1111/j.1095-8649.2012.03464.x, 
paleoDB 10.1111/ecog.01154,  rAvis 10.1371/journal.pone.0091650)}, meaning data 
associated with taxa in a given phylogeny can be obtained directly in R. 
These packages, combined the language's support for literatre programming 
\hl{(Knuth\_1984, Xie\_2014)} make R an ideal platform for reproducable research
in phylogentics and comparative biology, as they allow a complete record of the 
steps taken in gathering, processing and analysing a given data set to be
produced.

Here we present \texttt{rotl}, and R package that allows users to download
phylogentic and taxonomic data from Open Tree of Life in R. \texttt{rotl} takes
adavantage of Application Programming Interfaces (APIs) provided by OTL to acess
sub-trees from the synthetic Open Tree and the published trees that contribute
to the synthesis. By providing direct access to high quality
phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
analysis workflows, and extends the degree to which R supports reproducable
research in phylogenetics and comparative biology.

@@

In recent years, R has become part of the toolbox of many researchers in
evolutionary biology. R greatly facilitates the analysis of large datasets, and
allows researchers to combine methods in novel ways because many methods for
comparative analyses are implemented, and because it is a relatively easy to use
programming language. Additionally, as more data is made available online and
accessible using APIs, several packages have been developed to interact and
download these datasets directly in R enabling direct and reproducible analyses.
Notably, the organization rOpenSci \url{https://ropensci.org} has fostered a
community of researchers who develop tools and methods to facilitate the use of
open data and broaden the adoption of open science practices in
general. For instance, the rOpenSci-developed pacakge \texttt{rTreeBase} allows
users to acess phylogenies stored in treebase. \texttt{rotl} contributes to this
initiative, and greatly extends the number of taxa for which phylogenetic data
can be retreived within R while allowing the  data from OTL can be combined with 
other sources easily.

The package vignettes provide two demonstrations of the integration of a
phylogeny and data associated the taxa it represents. Specifically, the 
``Data mashups'' vignette provides an example of how data associated with tip-taxa
can be gathered and visualized. Another vigenette titled ``meta-analysis'' 
demonstrates how a complete analysis, including the gathering of data and a
phylogeny, can be performed in a single R session. We further demonstrate the
integration of phylogenetic data retrieved with \texttt{rotl} with other other
data here. We how how we can obtain a map of the occurrences for some of the cat species
(genus \textit{Lynx}) included in the phylogeny above using records for these
species found in GBIF (code available in Appendix)

>>>>

The R programming language is a popular tool for  phylogenetics and comparative
analysis. The R packages \texttt{ape} \cite{Paradis2004}, \texttt{phylobase}
\cite{phylobase-0.8.0 }and \texttt{RNeXML} \cite{Boettiger2015} provide
functions to import and manipulate trees within R and save the results in
standard data formats. Additional packages including  \texttt{phytools}
\cite{Revell2012}, \texttt{geiger} \cite{Pennell2014-geiger} and \texttt{ggtree} 
\cite{Yu-ggtree} to allow users to analyse and visualise data in a 
phylogenetic context (see 
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list of phylogenetics packages in R). In addition to packages for
phylogentic analyses, a growing number of R packages allow users to query and
acess data from the web \hl{(rFISHBASE 10.1111/j.1095-8649.2012.03464.x, 
paleoDB 10.1111/ecog.01154,  rAvis 10.1371/journal.pone.0091650)}, meaning data 
associated with taxa in a given phylogeny can be obtained directly in R. 
These packages, combined the language's support for literature programming 
\hl{(Knuth\_1984, Xie\_2014)} make R an ideal platform for reproducible research
in phylogenetics and comparative biology, as they allow a complete record of the 
steps taken in gathering, processing and analysing a given data set to be
produced.

Here we present \texttt{rotl}, and R package that allows users to download
phylogenetic and taxonomic data from Open Tree of Life in R. \texttt{rotl} takes
advantage of Application Programming Interfaces (APIs) provided by OTL to acess
sub-trees from the synthetic Open Tree and the published trees that contribute
to the synthesis. By providing direct access to high quality
phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
analysis workflows, and extends the degree to which R supports reproducable
research in phylogenetics and comparative biology.

@@

In recent years, R has become part of the toolbox of many researchers in
evolutionary biology. R greatly facilitates the analysis of large datasets, and
allows researchers to combine methods in novel ways because many methods for
comparative analyses are implemented, and because it is a relatively easy to use
programming language. Additionally, as more data is made available online and
accessible using APIs, several packages have been developed to interact and
download these datasets directly in R enabling direct and reproducible analyses.
Notably, the organization rOpenSci \url{https://ropensci.org} has fostered a
community of researchers who develop tools and methods to facilitate the use of
open data and broaden the adoption of open science practices in
general. For instance, the rOpenSci-developed pacakge \texttt{rTreeBase} allows
users to access phylogenies stored in treebase. \texttt{rotl} contributes to this
initiative, and greatly extends the number of taxa for which phylogenetic data
can be retrieved within R while allowing the  data from OTL can be combined with 
other sources easily.

The package vignettes provide two demonstrations of the integration of a
phylogeny and data associated the taxa it represents. Specifically, the 
``Data mashups'' vignette provides an example of how data associated with tip-taxa
can be gathered and visualized. Another vignette titled ``meta-analysis'' 
demonstrates how a complete analysis, including the gathering of data and a
phylogeny, can be performed in a single R session. We further demonstrate the
integration of phylogenetic data retrieved with \texttt{rotl} with other other
data here. We how how we can obtain a map of the occurrences for some of the cat species
(genus \textit{Lynx}) included in the phylogeny above using records for these
species found in GBIF (code available in Appendix)
